Group polarization replications may be marred by high false discovery rates

### 2024-12-19 GroupPolarizationStatmod/paper/ms.tex
all materials ready for final draft

\usepackage{amsmath,amssymb}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{xcolor}
\usepackage[margin=1.1in,letterpaper]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[sorting=none, style=numeric-comp]{biblatex}
\addbibresource{/Users/mt/workspace/Writing/library.bib}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Foundational claims of group polarization are plausibly false, with lessons for the empirical study of psychological dynamics},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

@@

\usepackage{authblk}
\usepackage{etoolbox}
\makeatletter
\title{Decades of ``group polarization'' replications may be undermined by high false discovery rates}
% \title{Foundational claims of group polarization are plausibly false,
% with lessons for the empirical study of psychological dynamics}
\makeatother
\author[1,*]{{Matthew A.~Turner}}
\affil[1]{\small Division of Social Sciences, Stanford Doerr School of Sustainability, Stanford University}

@@

\date{}
\begin{document}
\maketitle
\begin{abstract}
\noindent 
\emph{Group polarization} is the name for a form of consensus where
members of a like-minded group become, on average, more extreme in their
opinions after discussing a topic. Group polarization is important because
it may increase social and political tensions if moderates become more
extreme. Decades of studies have replicated detections of group
polarization, typically using an ordinal scale to measure opinions.
However, these studies did not account for ordinal measurements, which can
result in spurious detections of opinion change.  Lacking original data we
can still calculate the probability that a published detection of group
polarization was spurious, i.e., to calculate the \emph{false detection
rate}.  Across 54 group polarization experimental conditions from ten
representative journal articles, we calculated false detection rates to be
between 0.52 and 0.88, with a median of 0.75, using a generative model of
group polarization experiments seeded with data from the studies
themselves.  We also use our model to develop experimental designs that
achieve an acceptable false discovery rate.  Much of group polarization
research may be unreliable, but this analysis can help change that. More
broadly, this work demonstrates one way that replication success alone
does not imply epistemic reliability.
\end{abstract}

@@

\begin{figure}
  \caption{\textbf{Common components of a group polarization experiment.}
  Participants first give their opinion in response to some prompt (A). Afterwards,
  participants deliberate the topic (B). Participants then give their opinion again
  (C). Opinions are implicitly behavioral choices that are most consistent with some
  complex aggregation of brain states and dynamics, and so are theoretically
  continuous (D).}
  \centering
    \includegraphics[width=0.7\textwidth]{Figures/ModelSketchesA-C.pdf}
    \includegraphics[width=0.65\textwidth]{Figures/ModelSketchesD.pdf}
  \label{fig:ModelSketch}
\end{figure}

@@

@@

\subsection{Metascience of the false detection rate}

@@

We calculate $\alpha = \Pr(D | \neg E)$ through simulation for experimental
conditions identified as plausibly generating false detections, 
and use generously optimistic values of
$b=0.1$~\cite{Ioannidis2005,McElreath2015} and $W=0.8$~\cite{Smaldino2016NatSel}
in the main text and in the compare with more pessimistic, yet realistic, estimates
of these values, as well as even more optimistic estimates.

@@

@@

First, we assume that participant $i$ has
latent opinion at time $t$ that is drawn from a group-level normal distribution,
\begin{equation}
  o_{i,t} \sim p(o; \mu_t, \sigma_t) = \mathcal{N}(\mu_t, \sigma_t).
  \label{eq:opinionDistribution}
\end{equation} 
\noindent 
The initial opinion of agent $i$ is are denoted by $o_{i,t=0}$, and thus
in the model are drawn from a normal distribution with mean $\mu_0$ and standard
deviation $\sigma_0$. The final opinion of agent $i$ is written $o_{i,t=T}$,
drawn from a normal distribution with mean $\mu_T$ and standard deviation
$\sigma_T$.

\subsection{Root-finding algorithm to identify plausible false detections}

Simple consensus can appear as group polarization since a change in

\begin{algorithm}
  \caption{User-guided routine to find pre- and post-deliberation variances that
  yield group polarization in the absence of mean opinion change.
\mt{NOT YET DONE...}}\label{alg:routine}
  \begin{algorithmic}
    \Require $\mu$, $\sigmapre$, $\sigmapost$ \Comment{Static guess for common mean,
    initial guesses of variances.}
    \Require $\epsilon$, $\delta$, $i_\mathrm{max}$ \Comment{Root-finding tolerance, step size, and maximum
    iterations.}
    % \Require $\delta$ \Comment{Root-finding step size.}
    % \Require $i_\mathrm{max}$ \Comment{Maximum number of iterations.}
    \State $\Delta \gets \mupre' - \mupost'$ 
    \While {$\epsilon < |\Delta| $ \and $i < i_max$}
      \State $\Delta \gets (\mupre' - \mupost')^2$
    \EndWhile
  \end{algorithmic}
\end{algorithm}

>>>>

\usepackage{amsmath,amssymb}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{iftex}
\input{preamble.tex}
\hypersetup{
  pdftitle={Group polarization replications may be marred by high false discovery rates},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue}}

@@

\newcommand{\meanobs}{\bar{o}}
\newcommand{\meanobsemp}{\bar{\mathbf{o}}}
\newcommand{\meanobst}{\meanobs_t}
\newcommand{\meanobsjt}{\meanobs_{j,t}}

\newcommand{\meanobspre}{\meanobs_\mathrm{pre}}
\newcommand{\meanobspost}{\meanobs_\mathrm{post}}

@@

\usepackage{authblk}
\usepackage{etoolbox}
\usepackage{wrapfig}
\makeatletter
\title{Group polarization replications may be marred by high false discovery rates}
\makeatother
\author[1,*]{{Matthew A.~Turner}}
\affil[1]{\small Division of Social Sciences, Stanford Doerr School of Sustainability, Stanford University}

@@

\date{}
\begin{document}
\maketitle
\begin{abstract}
\noindent 
\emph{Group polarization} is the name for a form of consensus where
members of a like-minded group become, on average, more extreme in their
opinions after discussing a topic. Group polarization is important because
it may increase social and political tensions if moderates become more
extreme. Decades of studies have replicated detections of group
polarization, typically using an ordinal scale to measure opinions.
However, these studies did not account for ordinal measurements, which can
result in spurious detections of opinion change.  Lacking original data we
can still calculate the probability that a published detection of group
polarization was spurious, i.e., to calculate the \emph{false detection
rate}.  Across 54 group polarization experimental conditions from ten
representative journal articles, we calculated false detection rates to be
between 0.52 and 0.88, with a median of 0.75, using a generative model of
group polarization experiments seeded with empirical data.  
We also use our model to develop experimental designs that
achieve an acceptable false discovery rate.  Much of group polarization
research may be unreliable. This analysis can help change that by enabling others to avoid this
hidden pitfall. More
broadly this work demonstrates one important way that replication success alone
does not imply epistemic reliability.
\end{abstract}

@@

% \begin{figure}
%   \caption{\textbf{Opinions are implicitly behavioral choices that are most consistent with some
%   complex aggregation of brain states and dynamics, and so are theoretically
%   continuous (A). Common components of a group polarization experiment.}
%   Participants first give their opinion in response to some prompt (B). Afterwards,
%   participants deliberate the topic (C). Participants then give their opinion again
%   (D). }
%   \centering
%     \includegraphics[width=0.65\textwidth]{Figures/Model/ModelSketchesA.pdf}
%     \includegraphics[width=0.7\textwidth]{Figures/Model/ModelSketchesB-D.pdf}
%   \label{fig:ModelSketch}
% \end{figure}

@@

% \begin{wrapfigure}{l}{0.5\textwidth}
\begin{figure}
  \caption{\textbf{Group polarization experimental model.} Latent opinions are 
    represented as numbers corresponding to the valence and extremism of one's opinion
(A). Before deliberating a topic of interest, all participants report their 
opinion (B). Participants then deliberate the topic with others (C). Following
deliberation participants report their opinion once more (D).}
  \centering
  \includegraphics[width=0.65\textwidth]{Figures/Model/ModelSketches.pdf}
  \label{fig:experiment_model}
% \end{wrapfigure}
\end{figure}

@@

\subsection{Metascience of the false detection rate in group polarization}

@@

@@

The group polarization studies are journal articles that present
the results of multiple experiments to support some hypothesis. To organize
the presentation of false detection rates across the ten studies in our
analysis, we first will organize false detection rates by experiment, where
we created `ExperimentID` tags for each of the sixty experiments across ten
studies. We also created `StudyID` tags for each of the ten journal articles
with the key `{Author}{Year}`, e.g., \texttt{Schkade2010} for Schkade, et al.,
(2010) \cite{Schkade2010}.

We calculate $\alpha = \Pr(D | \neg E)$ through simulation for experimental
conditions, 
and use optimistic values of
$b=0.1$~\cite{Ioannidis2005,McElreath2015} and $W=0.8$~\cite{Smaldino2016NatSel}
in the main text and in the compare with more pessimistic, yet realistic, estimates
of these values, as well as even more optimistic estimates.

@@

An opinion is never measured directly, only a behavior that indicates an opinion
in a behavioral experiment or survey.  For example, the location of a mark
that a participant makes is measured, not the complex neural activity that
underlies forming opinions (Figure~\ref{fig:experiment_model}A) \mt{Some of this
needs to be Intro'd}.
Theoretically, an internal opinion in the form of neural activity is a
\textit{latent opinion}, from the Latin \emph{latentem} meaning ``lie
hidden''.  Participants report \text{observed opinions}, typically as one of
a finite number of ranked, successive integer values that form the
\emph{selection set}. The minimum possible value in the selection set might
mean ``strongly disagree'' and the maximum value represents ``strongly
agree''.

% \begin{wrapfigure}{r}{0.5\textwidth}
\begin{figure}
  \caption{\textbf{Opinion measurement model and spurious group polarization.} 
  Latent model participant opinions are drawn from a normal distribution with the
  listed parameters (A,C). The pre-deliberation latent distribution has high
  variance, but appears polarized when measured via integration (A).  The latent
  distribution's variance is expected to decrease during deliberation (B) via
  consensus, but in this example the latent mean is static, 
$\mupre = \mupost$. However, when the post-deliberation distribution is measured,
the simulated observed mean increases, i.e., spurious
group polarization occurs (C).}
  \centering
  \includegraphics[width=0.65\textwidth]{Figures/Model/latent-ordinal-distros.pdf}
  \label{fig:distros}
\end{figure}
% \end{wrapfigure}

Group polarization experiments share a common experimental design where,
first, participant opinions are gathered on a topic of interest. Next, groups
are formed that have some initial opinion extremism. Participants deliberate
or discuss the topic for some set time. After the set time, participants
again give their opinion. Group polarization is said to occur if the
mean group opinion became more extreme, i.e., if the magnitude of the mean
group opinion is greater \emph{post-deliberation} compared to
\emph{pre-deliberation}. For example,
\cite{Moscovici1969} asked Parisian high school students to indicate their
degree of agreement with the statement, ``American economic aid is always
used for political pressure'' (Figure~\ref{fig:experiment_model}A). There
were seven options in the selection set, -3 indicating ``strongly disagree'',
+3 indicating ``strongly agree'', and 0 was neutral. Moscovici and Zavalloni
report that the mean pre-deliberation opinion was -0.61 and the mean
post-deliberation opinion was -1.09 (Moscovici and Zavalloni, 1969; Table 4,
p. 132). This is a -0.48 shift to more extreme disagreement with the 
statement, reported by Moscovici and Zavalloni as statistically significant
in Table 5 on p. 132.

We formalize this experimental design in the $N\to\infty$ case where the
distribution of ordinal measurements is created by integrating continuous latent
opinion distributions.  Latent opinions in the model are normally distributed with
mean $\mu_t$ and variance $\sigma_t$ for time $t$. We are concerned with only two
times, \emph{pre-} and \emph{post-} deliberation ($t=0$ and $t=T$, respectively).
For short we can write $\mu_\mathrm{pre}$ and $\sigma_\mathrm{pre}$ for
pre-deliberation mean and variance (and similar for \emph{post}-deliberation).  
The latent opinion of participant $i$ at time $t$ is drawn from a normal 
distribution in the model, 
\begin{equation} 
  o_{i,t} \sim p(o; \mu_t, \sigma_t) = \mathcal{N}(\mu_t, \sigma_t).
  \label{eq:opinionDistribution} 
\end{equation} 
\noindent 
Latent opinions are assumed to be normally distributed since they are unbounded 
aggregations of neurobiological activity.

The measurement process with $N\to\infty$ participants is simulated by 
integrating the latent opinion distribution between ordinal bin thresholds, 
of which there are $B+1$, where $B$ is the number of opinion bins, with sequential
integer values $b$. In the Moscovici and Zavalloni (1969) example, $b \in
\{-3,-2,\ldots,2,3\}$ and $B = 7$.
Bin $b$ is defined by its lower and upper continuous-valued thresholds, 
$\theta_{b-1}$ and
$\theta_{b}$, respectively. Thresholds are given by the following equations
(and visualized as dotted lines in Figure~\ref{fig:distros}).
\begin{equation}
\begin{cases}
  \theta_0 &= -\infty \\
  \theta_{B} &= \infty \\
  \theta_b &= b + \frac{1}{2} \text{ if } 0 < b < B.
\end{cases}
\end{equation}
\noindent
With thresholds defined, the frequency of ordinal measurment $b$ at time $t$ 
is (Figure~\ref{fig:distros})
\begin{equation}
  p(b;~\mu_t, \sigma_t, B) = \int_{\theta_{b-1}}^{\theta_b} p(o; \mu_t, \sigma_t)
  do.
  \label{eq:ordinal-frequency}
\end{equation}

We can now define spurious group polarization and give examples of when it occurs.
First, the observed group polarization opinion shift is
\begin{equation}
  s = \meanobs_\mathrm{post} - \meanobs_\mathrm{pre}~,
  \label{eq:group-polarization}
\end{equation}
\noindent
where the expected value of observed opinions at time $t$ is
\begin{equation}
  \meanobst = \frac{1}{B} \sum_{b=1}^B b \cdot p(b; \mu_t, \sigma_t, B).
  \label{eq:meanobs}
\end{equation}
\noindent
\emph{Spurious} group polarization is when mean observed opinions seem to increase in 
extremity when the latent mean held constant. We simulate spurious
group polarization by setting $\mupre = \mupost = \mu$ and setting the variances
to values known to induce spurious group polarization. 
We explain how to find a constant latent mean, $\mu$, and pre- and post-deliberation
variances, $\sigmapre$ and $\sigmapost$, respectively, that generate spurious 
detections of group polarization in the next section. 
% We describe how we find $\mu$,
% $\sigmapre$ and $\sigmapost$ next in Section~\ref{sec:finding-variances}.

\subsection{Web app controls algorithm to find parameters inducing spurious group polarization}

To calculate the false discovery rate for each experimental condition, we first must
identify a constant latent mean, $\mu$, and latent variances $\sigmapre$ and
$\sigmapost$ that generate observation distributions\mt{define earlier?} with
reported mean pre- and post-deliberation opinions and resulting shift $s$ (calculated
using Equation~\ref{eq:group-polarization}). Define the collection of these
parameters $\beta = \{\mu, \sigmapre, \sigmapost\}$. We identify $\beta$ through a
hillclimbing optimization algorithm that finds $\beta$ that minimizes the difference
between reported pre- and post-deliberation mean opinions and simulated ones
calculated with Equation~\ref{eq:meanobs}.

The algorithm to find $\beta$ holds $\mu$ constant while searching for $\sigmapre$
and $\sigmapost$ that generate observed $\meanobspre$ and $\meanobspost$, within a
tolerance of $\pm10^{-3}$ (Algorithm~\ref{alg:hillclimbing}). 
The algorithm requires a researcher to specify an initial
guess for $\beta$ and auxiliary parameter settings, which generally vary between
experimental conditions.  Furthermore, the algorithm does not always produce
reasonable distributions with a given $\beta$, and sometimes it does not produce
spurious group polarization.  Therefore, it is necessary to sanity check the latent
and ordinal opinion distributions for the $\beta$ found by the algorithm. If all is
well, the experimental condition can be marked as plausibly spurious.  To tame the
complexity involved in data entry and model fitting for each of these sixty
experimental conditions, we developed a web application to input data from control
the algorithm and inspect its results (Figure~\ref{fig:webapp}). Web application code is
available online in our GitHub
repository\footnote{\url{https://github.com/mt-digital/gp-statmod/blob/main/GroupPolarizationStatmod/app.R}}.

\begin{algorithm}
  \caption{Routine to find latent mean and variance that generate spurious group polarization.}
  \label{alg:hillclimbing}
  \begin{algorithmic}

    \Require $\mu$, $\sigma$, $\bar{\mathbf{o}}$ 
      \Comment{Static guess for common mean, initial guess of variance.}

    \Require $\epsilon$, $\delta$, $i_\mathrm{max}$ 
      \Comment{Hillclimbing tolerance, step size, and maximum iterations.}

    \State $\Delta \gets \meanobs(\mu, \sigma) - \meanobsemp$ 
      \Comment{Initialize simulation error.}
    \While {$\Delta^2 > \epsilon$ and $i < i_\mathrm{max}$} \Comment{Search until desired
    accuracy or max iterations reached.}
      \State $\sigma' \sim \mathcal{N}(\sigma, \delta)$ \Comment{Draw new variance to test.}
      \State $\Delta' \gets \bar{o}(\mu, \sigma') - \bar{\mathbf{o}}$
      \If {$\Delta' - \Delta < 0$}
        \State $\sigma \gets \sigma'$
          \Comment{Update $\sigma$ if the error is less than before.}
      \EndIf
      \State $\Delta \gets \bar{o}(\mu, \sigma) - \bar{\mathbf{o}}$
        \Comment{Update simulation error if $\sigma$ is updated.}
    \EndWhile \\

    \Return $\sigma$
  \end{algorithmic}
\end{algorithm}

\begin{figure}
  \caption{\textbf{Shiny web application for user-guided hillclimbing to find
  parameters that generate spurious group polarization.} This web application enables users to
create new analyses of experimental conditions from published studies. Users 
provide initial guesses for the algorithm to use as starting points in 
finding variances that generate spurious group polarization. Users inspect 
model fits provided by the algorithm, adjusting parameters if necessary, and
mark an experimental condition as ``Plausibly spurious'' if a reasonable fit
is indeed found.\mt{Maybe add some A, B, C guides to the figure and include
in parentheses here to go with the text?}}
  \centering
    \includegraphics[width=0.75\textwidth]{Figures/Model/WebApp.png}
  \label{fig:webapp}
\end{figure}

Note that so far, we have only identified cases where spurious group polarization plausibly
occurs in the asymptotic case where $N\to\infty$ ($N$ being the number of participants).
Experimental conditions yielding plausibly spurious group polarization warrant 
some suspicion. If there exists set of latent simple consensus 
parameters $\beta$ that create binned (Equation~\ref{eq:ordinal-frequency})
opinion distributions that naively seem to be group polarized, then the means
of at least some samples from these distibutions with empirical numbers of 
participants would also appear to be shifted consistent with group
polarization. This process itself does not tell us the false detection rate.
We use the identified simple consensus parameters to define pre- and post-deliberation
latent opinion distributions in simulated trials of group polarization 
experiments.
